Research Design 2:\\ Comparing Reinforcement Learning and Finite State Machine Agents in Real Time Strategy Games: Impact on Player Experience

### 2025-05-25 Paper/main.tex
copied over the findings and discussion part, and added the dataset repo as well as the zotero export.

% Rewrite parts of this, since it was AI formatted, and may contain errors
\section{Findings and Discussion}

\subsection{Chapter Overview}
This section presents the results of the playtest experiment, including quantitative and qualitative findings, and discusses their implications in the context of the research questions and literature.

\subsection{Quantitative and Qualitative Results with Discussion}

A total of 20 participants played three sessions each (Easy, Medium, Hard difficulty) against both FSM and RL agents, resulting in 120 valid game sessions.
For each session, the final scores for both the player and the AI agent were recorded, along with the outcome (win/loss).
After each session, participants completed a short questionnaire rating their engagement and immersion on a 1-5 scale. The results are summarized in Table~\ref{tab:results}.

\begin{table}[h]
\centering
\caption{Average Player and Agent Scores, Win Rate, Engagement, and Immersion by Difficulty}
\label{tab:results}
\begin{tabular}{lcccccc}
\hline
\multirow{2}{*}{\textbf{Metric}} & \multicolumn{2}{c}{\textbf{Easy}} & \multicolumn{2}{c}{\textbf{Medium}} & \multicolumn{2}{c}{\textbf{Hard}} \\
 & FSM & RL & FSM & RL & FSM & RL \\
\hline
Player Avg. Score         & 470 & 489 & 540 & 525 & 584 & 560 \\
Agent Avg. Score          & 390 & 364 & 413 & 460 & 433 & 520 \\
Player Win Rate           & 95\% & 100\% & 90\% & 85\% & 85\% & 70\% \\
Avg. Score Diff. (\%)     & 17.0\% & 24.4\% & 22.4\% & 12.4\% & 25.5\% & 7.1\% \\
Engagement (1-5)          & 3.0 & 2.9 & 3.2 & 3.8 & 3.3 & 4.2 \\
Immersion (1-5)           & 2.9 & 2.8 & 3.0 & 3.7 & 3.1 & 4.1 \\
\hline
\end{tabular}
\end{table}

\textbf{Easy Difficulty:}  
FSM agents provided a greater challenge at the easy level, with players winning all games but with a higher average score difference compared to RL. RL agents were slightly easier at this level,
with players maintaining a high win rate and reporting slightly higher engagement and immersion than with FSMs. This suggests that, at lower difficulty, FSMs can still provide a meaningful challenge,
while RL agents may be less aggressive or adaptive at this stage.

\textbf{Medium Difficulty:}  
As difficulty increased, RL agents became more competitive, narrowing the score gap and reducing the player win rate. Engagement and immersion scores for RL agents increased more sharply than for FSMs,
reflecting the RL agent's adaptability and unpredictability. This aligns with literature suggesting RL agents can better match player skill as difficulty rises \cite{grech_creating_2023}.

\textbf{Hard Difficulty:}  
At the hardest level, RL agents surpassed FSMs in challenge,with the agent's average score nearly matching the player's and the player win rate dropping to 70\%.
Engagement and immersion were highest for RL agents at this level, indicating that players found the experience both challenging and captivating.
FSM agents, while still providing a challenge, were less able to keep up with player performance at higher difficulty.
This supports the hypothesis that RL agents can provide a more engaging and immersive experience, especially at higher difficulty levels \cite{vinyals_grandmaster_2019}.

\textbf{Qualitative Feedback:}  
Open-ended feedback revealed that FSM agents were often described as predictable, especially at higher difficulties:  
\textit{“After a few rounds, I could predict what the FSM AI would do, even on hard.”}

In contrast, RL agents were seen as increasingly challenging and engaging as difficulty increased:  
\textit{“The RL agent felt more like playing against a real person, especially on hard mode, but sometimes it was a bit too hard.”}

Some participants noted that the RL agent's unpredictability increased immersion, but a few found it frustrating when the difficulty spiked.
Several participants also commented that they felt more engaged and “in the zone” when playing against the RL agent, particularly at higher difficulties.

\subsection{Chapter Overview}
Overall, the combination of quantitative scores and qualitative feedback indicates that RL agents can enhance both engagement and immersion for players,
but care must be taken to ensure the challenge remains fair and enjoyable across all difficulty levels.
The results highlight the importance of difficulty balancing for RL agents, as excessive challenge can lead to frustration,
while FSM agents may fail to sustain long-term engagement due to their predictability.

% \section{Conclusion}

% This study aimed to compare the impact of RL and FSM agents on player experience in RTS games, focusing on engagement, immersion, and perceived difficulty. By implementing both AI approaches in a custom RTS prototype,
% a controlled playtest was conducted with 30 participants across three difficulty levels. The results demonstrated that RL agents, particularly at higher difficulties, provided a more engaging and immersive experience
% for players compared to FSM agents. Contrarily, FSM agents were found to provide a more meaningful challenge at lower levels, and were easier to balance. This predictability however, was the reason for reduced
% player engagement and immersion at higher difficulties, which is in direct contrast to RL agents, where immersion and engagement increased with difficulty.

% These findings support the hypothesis that RL agents can enhance player experience in RTS games, specifically on harder difficulties, which could justify their higher development and computational costs in contexts
% where this increased difficulty is desired. However, for the easier difficulties, and general scenarios, FSM agents remain the better option, due to the ease of implementation, balancing, and low computational cost.

@@

% TODO: discuss the balancing issues further in the findings and the conclusion, as most balance changes required a retraining of the RL agent, where as the FSM was either left unchanged, or slightly tweaked.

@@

% TODO: Finalise the changes being done in the copy outside the repo, and place them here, since almost every other part is finalised.

>>>>

\section{Findings and Discussion}

\subsection{Chapter Overview}
Following the prototype testing, the data collected from the playtest sessions was analysed to evaluate the impact of RL and FSM agents on player experience.
The analysis focused on comparing quantitative metrics such as player enjoyment, engagement, and immersion, and aligning them with the qualitative feedback as well as the in-game analytics.
This was done as to provide a more comprehensive understanding of player experience, and to identify the key factors that influence it.

\subsection{Findings}

Below are the average results from the playtest sessions, comparing player performance against both FSM and RL agents across three difficulty levels: easy, medium, and hard.

\begin{table}[h]
	\centering
	\caption{Average Player and Agent Scores, Win Rate, Engagement, and Immersion by Difficulty}
	\label{tab:results}
	\begin{tabular}{lcccccc}
		\hline
		\multirow{2}{*}{\textbf{Metric}} & \multicolumn{2}{c}{\textbf{Easy}} & \multicolumn{2}{c}{\textbf{Medium}} & \multicolumn{2}{c}{\textbf{Hard}}                           \\
		                                 & FSM                               & RL                                  & FSM                               & RL     & FSM    & RL    \\
		\hline
		Player Avg. Score                & 470                               & 489                                 & 540                               & 525    & 584    & 560   \\
		Agent Avg. Score                 & 390                               & 364                                 & 413                               & 460    & 433    & 520   \\
		Player Win Rate                  & 95\%                              & 100\%                               & 90\%                              & 85\%   & 85\%   & 70\%  \\
		Avg. Score Diff. (\%)            & 17.0\%                            & 24.4\%                              & 22.4\%                            & 12.4\% & 25.5\% & 7.1\% \\
		Enjoyment (1-5)                  & 3.9                               & 3.7                                 & 4.1                               & 4.2    & 4.0    & 4.5   \\
		Engagement (1-5)                 & 3.0                               & 2.9                                 & 3.2                               & 3.8    & 3.3    & 4.2   \\
		Immersion (1-5)                  & 2.9                               & 2.8                                 & 3.0                               & 3.7    & 3.1    & 4.1   \\
		\hline
	\end{tabular}
\end{table}

\subsection{Discussion of Results}

As we can see in Table \ref{tab:results}, on the easiest difficulty, players had a higher average score and win-rate against RL agents, with a win rate of 100\%, compared to 95\% against FSM agents.
In medium and hard difficulties however, RL agents were more competitive, with the average score difference decreasing significantly, and the player win rate dropping to 85\% and 70\% respectively.
This indicates that the split in difficulty levels was greater for RL agents, as the easy difficulty was easier, and the hard difficulty was harder, compared to FSM opponents.

We can also see that the score difference while higher for RL agents at easy difficulty, was a lot lower on the higher difficulties. This indicates that RL agents were able to adapt better,
and remained more consistent in their performance, while FSM agents were more predictable and less adaptive.

Regarding player enjoyment, we can see that like with the RL performance, players enjoyed FSMs more at the easy difficulty, while RL agents were more enjoyable at the medium and hard difficulties.
The same trend happens with engagement and immersion, with RL agents achieving an even greater increase in engagement and immersion compared to FSMs on the medium and hard difficulties,
with the gap in easy difficulty being smaller.

This suggests that the tougher the opponent, the more engaging and immersive the experience became, aligning with the findings of \cite{grech_creating_2023} and \cite{vinyals_grandmaster_2019}.
As for enjoyment, while it did increase with the harder difficulties, likely due to it being more engaging and immersive, it was not as pronounced as the other two metrics, indicating that the increased
loss rate and difficulty may have led to some frustration for players, especially with RL agents at the hard difficulty.

This aligns pretty well with the qualitative feedback gathered from the participants, where most players reported that the increased challenge and adaptability of RL agents made the game more engaging and fun,
with a few mentioning that the increased difficulty sometimes made it feel unfair. They compared this to feeling like an unbalanced match, where the skill difference was too great.
On the hard difficulty especially, a lot of feedback was given that the FSM agents were becoming easier, as while they were faster and making smarter decisions, it was clear what logic they were following,
and as such, some of the players were able to predict their actions and exploit them. This could help explain the higher average score difference for FSM agents at the hard difficulty, as players had learned
how to exploit it.

When aligning this with the hypothesis, we can see that it is partially supported, as while players on average reported higher levels of enjoyment and immersion when playing against RL agents,
this was only true at the medium and hard difficulties, with the easy difficulty being more enjoyable against FSM agents. This could have been due to the FSM agents being new to the players, and as such they
had not yet learned how to exploit them. This aligns well with the average score difference in Table \ref{tab:results}, where the score difference for FSMs was lowest at the easy difficulty.

% TODO: Discuss the following:
% You could add a short paragraph explicitly discussing validity (e.g., sample size, playtest design), reliability (consistency of results), and generalizability (applicability to other games or genres). 
% You touch on these in your methodology, but a brief mention here would strengthen the critique.

\subsection{Summary of Findings}
From the analysis of the playtest data, we saw that on easy difficulty, FSM agents provided a more meaningful challenge, and a better player experience than RL agents.
This could have been due to players not having learned how to exploit the FSM agents yet, or the RL agent being too inconsitent at this difficulty.
As the difficulty increased, RL agents became more competitive, with the average score difference decreasing significantly, and player experience metrics such as engagement and immersion increasing.
Contrarily, FSMs became more predictable and easier to exploit, leading to a decrease in player engagement and immersion at higher difficulties.
This means that hypothesis is only partially supported, as while players reported higher levels of enjoyment and immersion when playing against RL agents, this was not the case at the easy difficulty.

@@

@@
